---
title: "GCTB Tutorial"
author: "Jian Zeng <j.zeng@uq.edu.au>"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

This tutorial demonstrates:

1. Loading 1000G genotype data
2. Simulating individual-level phenotypes
3. Running BayesS analysis using individual-level data
4. Generating GWAS summary statistics
5. Generating sparse LD matrix
6. Running SBayesS analysis using sparse LD matrix
7. Generating block LD matrix
8. Running SBayesR analysis using eigen-decomposed LD matrix
9. Running SBayesRC analysis with annotation for causal variants

# Requirements

- R `>= 4.0` with `BEDMatrix` package
- Compiled `gctb` binary at `scr/gctb`

# Load data

```{r}
library(BEDMatrix)

# Set paths
if (basename(getwd()) == "tutorial") {
  data_dir <- "data"
  result_dir <- "results"
  gctb <- "../scr/gctb"
} else {
  data_dir <- "tutorial/data"
  result_dir <- "tutorial/results"
  gctb <- "scr/gctb"
}

if (!dir.exists(result_dir)) dir.create(result_dir, recursive = TRUE)

# Load 1000G data
bed_file <- file.path(data_dir, "1000G_eur_chr22")
fam <- read.table(paste0(bed_file, ".fam"), stringsAsFactors = FALSE,
                  col.names = c("FID", "IID", "FATHER", "MOTHER", "SEX", "PHENO"))
bim_all <- read.table(paste0(bed_file, ".bim"), stringsAsFactors = FALSE,
                      col.names = c("CHR", "SNP", "CM", "BP", "ALLELE_A", "ALLELE_B"))

bed <- BEDMatrix(bed_file)
geno <- as.matrix(bed)

n_ind <- nrow(fam)
n_snp_all <- nrow(bim_all)
cat("Loaded:", n_ind, "individuals,", n_snp_all, "SNPs\n")

# Filter SNPs with MAF < 0.01
maf_all <- colMeans(geno, na.rm = TRUE) / 2
maf_all <- pmin(maf_all, 1 - maf_all)  # Get minor allele frequency
keep_snp <- maf_all >= 0.01

geno <- geno[, keep_snp]
bim <- bim_all[keep_snp, ]
n_snp <- sum(keep_snp)
cat("After MAF filtering (MAF >= 0.01):", n_snp, "SNPs\n")
```

# Simulate phenotypes

```{r}
set.seed(555)

# Simulation settings
h2 <- 0.5  # heritability
n_causal <- 10  # number of causal variants
S <- -1  # the S parameter

# Sample causal variants
causal_idx <- sample(n_snp, n_causal)

# Calculate allele frequencies for causal variants
freq_cv <- colMeans(geno[, causal_idx], na.rm = TRUE) / 2

# Sample causal variant effects
beta_cv <- rnorm(n_causal, mean = 0, sd = sqrt((2 * freq_cv * (1 - freq_cv))^S))

# Store true effects for all SNPs (zeros for non-causal)
beta_true <- rep(0, n_snp)
beta_true[causal_idx] <- beta_cv

# Calculate true genetic values
g <- geno[, causal_idx] %*% beta_cv

# Simulate phenotypes
e <- rnorm(n_ind, 0, sqrt(var(g, na.rm = TRUE) * (1 - h2) / h2))
y <- g + e

# Save phenotype file
phen_path <- file.path(data_dir, "sim.phen")
write.table(data.frame(FID = fam$FID, IID = fam$IID, PHENO = y),
            phen_path, quote = FALSE, sep = "\t", row.names = FALSE, col.names = FALSE)
```

# Run BayesS using individual-level data

```{r}
# Individual-level analysis
# Command line format (can be copied and run in terminal):
# gctb --bfile data/1000G_eur_chr22 --pheno data/sim.phen --bayes S --out results/bayess

system2(gctb, c(
  "--bfile", bed_file,
  "--pheno", phen_path,
  "--bayes", "S",
  "--out", file.path(result_dir, "bayess")
))

# Inspect results
snp_res <- read.table(file.path(result_dir, "bayess.snpRes"), header = TRUE)
par_res <- read.table(file.path(result_dir, "bayess.parRes"), header = TRUE)

cat("SNP results:\n")
head(snp_res)
cat("\nPosterior estimates of model parameters:\n")
print(par_res)

# Plot PIP values
pip_col <- if ("PIP" %in% colnames(snp_res)) "PIP" else "pip"
pip <- snp_res[[pip_col]]
name_col <- if ("Name" %in% colnames(snp_res)) "Name" else "SNP"
causal_snps <- bim$SNP[causal_idx]
is_causal <- snp_res[[name_col]] %in% causal_snps

plot(pip, type = "h", xlab = "SNP", ylab = "Posterior inclusion probability",
     col = "gray", lwd = 1)
points(which(is_causal), pip[is_causal], col = "red", pch = 19, cex = 1.2)
segments(which(is_causal), 0, which(is_causal), pip[is_causal], 
         col = "red", lwd = 2)
legend("topright", legend = c("Non-causal", "Causal"), 
       col = c("gray", "red"), lwd = c(1, 2), pch = c(NA, 19))
```

# Generate GWAS summary statistics

```{r}
# Calculate allele frequencies
af <- colMeans(geno, na.rm = TRUE) / 2

# Simple linear regression for each SNP
gwas_sumstats <- data.frame(
  SNP = bim$SNP,
  A1 = bim$ALLELE_B,
  A2 = bim$ALLELE_A,
  FREQ = af,
  BETA = NA,
  SE = NA,
  P = NA,
  N = n_ind
)

for (i in 1:n_snp) {
  fit <- lm(y ~ geno[, i])
  gwas_sumstats$BETA[i] <- coef(fit)[2]
  gwas_sumstats$SE[i] <- summary(fit)$coefficients[2, 2]
  gwas_sumstats$P[i] <- summary(fit)$coefficients[2, 4]
}

# Save summary statistics
sumstats_path <- file.path(data_dir, "sim.ma")
write.table(gwas_sumstats, sumstats_path, quote = FALSE, sep = "\t",
            row.names = FALSE, col.names = TRUE)
head(gwas_sumstats)
```

# Generate sparse LD matrix

```{r}
# Create sparse LD matrix from genotype data
# The sparse format stores only non-zero LD correlations within a window around each SNP
# Note: GCTB appends .ldm.sparse to the output path, so we use the base name
sparse_ldm_base <- file.path(data_dir, "1000G_eur_chr22")
sparse_ldm_path <- paste0(sparse_ldm_base, ".ldm.sparse")

# Command line format (can be copied and run in terminal):
# gctb --bfile data/1000G_eur_chr22 --make-sparse-ldm --out data/1000G_eur_chr22

system2(gctb, c(
  "--bfile", bed_file,
  "--make-sparse-ldm",
  "--out", sparse_ldm_base
))

# Check the output files
cat("\nSparse LD matrix files created:\n")
cat("  Info file:", paste0(sparse_ldm_path, ".info"), "\n")
cat("  Binary file:", paste0(sparse_ldm_path, ".bin"), "\n")
```

# Run SBayesS using sparse LD matrix

```{r}
# Summary-data-based analysis using sparse LD matrix
# This is more memory-efficient than using full LD matrices
# Command line format (can be copied and run in terminal):
# gctb --sbayes S --ldm data/1000G_eur_chr22.ldm.sparse --gwas-summary data/sim.ma --out results/sbayess

system2(gctb, c(
  "--sbayes", "S",
  "--ldm", sparse_ldm_path,
  "--gwas-summary", sumstats_path,
  "--out", file.path(result_dir, "sbayess")
))

# Inspect results
snp_res_sparse <- read.table(file.path(result_dir, "sbayess.snpRes"), header = TRUE)
par_res_sparse <- read.table(file.path(result_dir, "sbayess.parRes"), header = TRUE)

cat("SBayesS (sparse LDM) SNP results:\n")
head(snp_res_sparse)
cat("\nSBayesS (sparse LDM) posterior estimates of model parameters:\n")
print(par_res_sparse)
```

# Generate block LD matrix

```{r}
# Compute block LD matrix based on LD block map
# Note: You need to provide a block info file (e.g., ref_b37_1588blocks.pos)
block_info_file <- file.path(data_dir, "ref_b37_1588blocks.pos")

# Command line format (can be copied and run in terminal):
# gctb --bfile data/1000G_eur_chr22 --make-block-ldm --block-info data/ref_b37_1588blocks.pos --out data/ldm

system2(gctb, c(
  "--bfile", bed_file,
  "--make-block-ldm",
  "--block-info", block_info_file,
  "--out", file.path(data_dir, "ldm")
))

ldm_path <- file.path(data_dir, "ldm")

# Perform eigen-decomposition on the LD matrices with matched SNPs in GWAS data
# Command line format (can be copied and run in terminal):
# gctb --ldm data/ldm --gwas-summary data/sim.ma --make-ldm-eigen --out data/ldm

system2(gctb, c(
  "--ldm", ldm_path,
  "--gwas-summary", sumstats_path, 
  "--make-ldm-eigen",
  "--out", ldm_path
))
```

# Run SBayesR using eigen-decomposed matrices

```{r}
# Summary-data-based analysis
# Command line format (can be copied and run in terminal):
# gctb --gwas-summary data/sim.ma --ldm-eigen data/ldm --sbayes R --out results/sbayesr

system2(gctb, c(
  "--gwas-summary", sumstats_path,
  "--ldm-eigen", ldm_path,
  "--sbayes", "R",
  "--out", file.path(result_dir, "sbayesr")
))

# Inspect results
snp_res_sumstats <- read.table(file.path(result_dir, "sbayesr.snpRes"), header = TRUE)
par_res_sumstats <- read.table(file.path(result_dir, "sbayesr.parRes"), header = TRUE)

cat("SBayesR SNP results:\n")
head(snp_res_sumstats)
cat("\nSBayesR posterior estimates of model parameters:\n")
print(par_res_sumstats)

# Plot PIP values for SBayesR
pip_col <- if ("PIP" %in% colnames(snp_res_sumstats)) "PIP" else "pip"
pip <- snp_res_sumstats[[pip_col]]
name_col <- if ("Name" %in% colnames(snp_res_sumstats)) "Name" else "SNP"
causal_snps <- bim$SNP[causal_idx]
is_causal <- snp_res_sumstats[[name_col]] %in% causal_snps

plot(pip, type = "h", xlab = "SNP", ylab = "Posterior inclusion probability",
     col = "gray", lwd = 1, main = "SBayesR PIP Values")
points(which(is_causal), pip[is_causal], col = "red", pch = 19, cex = 1.2)
segments(which(is_causal), 0, which(is_causal), pip[is_causal], 
         col = "red", lwd = 2)
legend("topright", legend = c("Non-causal", "Causal"), 
       col = c("gray", "red"), lwd = c(1, 2), pch = c(NA, 19))
```

# Create annotation file for causal variants

```{r}
# Create annotation file with SNP, Intercept, and Anno columns for ALL SNPs in genotype file
# Anno = 1 if SNP is causal, 0 otherwise
# Map causal SNPs from filtered set to original set
causal_snps_filtered <- bim$SNP[causal_idx]  # Causal SNPs in filtered set
causal_snps_all <- bim_all$SNP[bim_all$SNP %in% causal_snps_filtered]  # Causal SNPs in original set

anno_data <- data.frame(
  SNP = bim_all$SNP,
  Intercept = 1,
  Anno = ifelse(bim_all$SNP %in% causal_snps_all, 1, 0)
)

anno_path <- file.path(data_dir, "causal_anno.txt")
write.table(anno_data, anno_path, quote = FALSE, sep = "\t", 
            row.names = FALSE, col.names = TRUE)

cat("Annotation file created for all SNPs in genotype file:\n")
cat("  Total SNPs:", nrow(anno_data), "\n")
cat("  Causal SNPs:", sum(anno_data$Anno), "\n")
cat("  Non-causal SNPs:", sum(anno_data$Anno == 0), "\n")
head(anno_data)
```

# Run SBayesRC with annotation

```{r}
# Summary-data-based analysis with annotation stratification
# Command line format (can be copied and run in terminal):
# gctb --gwas-summary data/sim.ma --ldm-eigen data/ldm --sbayes RC --annot data/causal_anno.txt --out results/sbayesrc

system2(gctb, c(
  "--gwas-summary", sumstats_path,
  "--ldm-eigen", ldm_path,
  "--sbayes", "RC",
  "--annot", anno_path,
  "--out", file.path(result_dir, "sbayesrc")
))

# Inspect results
snp_res_anno <- read.table(file.path(result_dir, "sbayesrc.snpRes"), header = TRUE)
par_res_anno <- read.table(file.path(result_dir, "sbayesrc.parRes"), header = TRUE)

cat("SBayesRC SNP results:\n")
head(snp_res_anno)
cat("\nSBayesRC posterior estimates of model parameters:\n")
print(par_res_anno)

# Plot PIP values for SBayesRC
pip_col <- if ("PIP" %in% colnames(snp_res_anno)) "PIP" else "pip"
pip <- snp_res_anno[[pip_col]]
name_col <- if ("Name" %in% colnames(snp_res_anno)) "Name" else "SNP"
causal_snps <- bim$SNP[causal_idx]
is_causal <- snp_res_anno[[name_col]] %in% causal_snps

plot(pip, type = "h", xlab = "SNP", ylab = "Posterior inclusion probability",
     col = "gray", lwd = 1, main = "SBayesRC PIP Values")
points(which(is_causal), pip[is_causal], col = "red", pch = 19, cex = 1.2)
segments(which(is_causal), 0, which(is_causal), pip[is_causal], 
         col = "red", lwd = 2)
legend("topright", legend = c("Non-causal", "Causal"), 
       col = c("gray", "red"), lwd = c(1, 2), pch = c(NA, 19))
```
