---
title: "GCTB Tutorial"
author: "Jian Zeng <j.zeng@uq.edu.au>"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

This tutorial demonstrates:

1. Loading 1000G genotype data
2. Simulating individual-level phenotypes
3. Generating GWAS summary statistics
4. Generating reference LD matrix (full and block)
5. Running BayesS analysis using individual-level data
6. Running SBayesS analysis using GWAS summary statistics

# Requirements

- R `>= 4.0` with `BEDMatrix` package
- Compiled `gctb` binary at `scr/gctb`

# Load data

```{r}
library(BEDMatrix)

# Set paths
if (basename(getwd()) == "tutorial") {
  data_dir <- "data"
  result_dir <- "results"
  gctb <- "../scr/gctb"
} else {
  data_dir <- "tutorial/data"
  result_dir <- "tutorial/results"
  gctb <- "scr/gctb"
}

if (!dir.exists(result_dir)) dir.create(result_dir, recursive = TRUE)

# Load 1000G data
bed_file <- file.path(data_dir, "1000G_eur_chr22")
fam <- read.table(paste0(bed_file, ".fam"), stringsAsFactors = FALSE,
                  col.names = c("FID", "IID", "FATHER", "MOTHER", "SEX", "PHENO"))
bim <- read.table(paste0(bed_file, ".bim"), stringsAsFactors = FALSE,
                  col.names = c("CHR", "SNP", "CM", "BP", "ALLELE_A", "ALLELE_B"))

bed <- BEDMatrix(bed_file)
geno <- as.matrix(bed)

n_ind <- nrow(fam)
n_snp <- nrow(bim)
cat("Loaded:", n_ind, "individuals,", n_snp, "SNPs\n")

# Filter SNPs with MAF < 0.01
maf_all <- colMeans(geno, na.rm = TRUE) / 2
maf_all <- pmin(maf_all, 1 - maf_all)  # Get minor allele frequency
keep_snp <- maf_all >= 0.01

geno <- geno[, keep_snp]
bim <- bim[keep_snp, ]
n_snp <- sum(keep_snp)
cat("After MAF filtering (MAF >= 0.01):", n_snp, "SNPs\n")
```

# Simulate phenotypes

```{r}
set.seed(123)

# Simulation settings
h2 <- 0.5  # heritability
n_causal <- 10  # number of causal variants
S <- -1  # the S parameter

# Sample causal variants
causal_idx <- sample(n_snp, n_causal)

# Calculate allele frequencies for causal variants
freq_cv <- colMeans(geno[, causal_idx], na.rm = TRUE) / 2

# Sample causal variant effects
beta_cv <- rnorm(n_causal, mean = 0, sd = sqrt((2 * freq_cv * (1 - freq_cv))^S))

# Store true effects for all SNPs (zeros for non-causal)
beta_true <- rep(0, n_snp)
beta_true[causal_idx] <- beta_cv

# Calculate true genetic values
g <- geno[, causal_idx] %*% beta_cv

# Simulate phenotypes
e <- rnorm(n_ind, 0, sqrt(var(g, na.rm = TRUE) * (1 - h2) / h2))
y <- g + e

# Save phenotype file
phen_path <- file.path(data_dir, "sim.phen")
write.table(data.frame(FID = fam$FID, IID = fam$IID, PHENO = y),
            phen_path, quote = FALSE, sep = "\t", row.names = FALSE, col.names = FALSE)
```

# Generate GWAS summary statistics

```{r}
# Calculate allele frequencies
af <- colMeans(geno, na.rm = TRUE) / 2

# Simple linear regression for each SNP
gwas_sumstats <- data.frame(
  SNP = bim$SNP,
  A1 = bim$ALLELE_B,
  A2 = bim$ALLELE_A,
  FREQ = af,
  BETA = NA,
  SE = NA,
  P = NA,
  N = n_ind
)

for (i in 1:n_snp) {
  fit <- lm(y ~ geno[, i])
  gwas_sumstats$BETA[i] <- coef(fit)[2]
  gwas_sumstats$SE[i] <- summary(fit)$coefficients[2, 2]
  gwas_sumstats$P[i] <- summary(fit)$coefficients[2, 4]
}

# Save summary statistics
sumstats_path <- file.path(data_dir, "sim.ma")
write.table(gwas_sumstats, sumstats_path, quote = FALSE, sep = "\t",
            row.names = FALSE, col.names = TRUE)
head(gwas_sumstats)
```

# Generate reference LD matrix

```{r}
# Generate full LD matrix
system2(gctb, c(
  "--bfile", bed_file,
  "--make-full-ldm",
  "--out", file.path(data_dir, "sim")
))

ldm_path <- file.path(data_dir, "sim.ldm.full")
```

# Generate block LD matrix

```{r}
# Compute block LD matrix based on LD block map
# Note: You need to provide a block info file (e.g., ref_b37_1588blocks.pos)
block_info_file <- "data/ref_b37_1588blocks.pos"  # Update with your block info file path

system2(gctb, c(
  "--bfile", bed_file,
  "--make-block-ldm",
  "--block-info", block_info_file,
  "--out", file.path(data_dir, "ldm")
))

ldm_path <- file.path(data_dir, "ldm")

# Perform eigen-decomposition on the LD matrices with matched SNPs in GWAS data
system2(gctb, c(
  "--ldm", ldm_path,
  "--gwas-summary", sumstats_path, 
  "--make-ldm-eigen",
  "--out", ldm_path
))
```

# Run BayesS using individual-level data

```{r}
# Individual-level analysis
system2(gctb, c(
  "--bfile", bed_file,
  "--pheno", phen_path,
  "--bayes", "S",
  "--out", file.path(result_dir, "sim")
))
```

# Inspect results

```{r}
snp_res_path <- file.path(result_dir, "sim.snpRes")
par_res_path <- file.path(result_dir, "sim.parRes")

if (file.exists(snp_res_path)) {
  snp_res <- read.table(snp_res_path, header = TRUE)
  cat("SNP results:\n")
  head(snp_res)
} else {
  message("Run GCTB first to create ", snp_res_path)
}

if (file.exists(par_res_path)) {
  par_res <- read.table(par_res_path, header = TRUE)
  cat("\nPosterior estimates of model parameters:\n")
  print(par_res)
} else {
  message("Run GCTB first to create ", par_res_path)
}

# Plot PIP values
if (file.exists(snp_res_path)) {
  # Read results
  snp_res <- read.table(snp_res_path, header = TRUE)
  
  # Find the PIP column
  pip_col <- NULL
  for (col in c("PIP", "pip")) {
    if (col %in% colnames(snp_res)) {
      pip_col <- col
      break
    }
  }
  
  if (!is.null(pip_col)) {
    pip <- snp_res[[pip_col]]
    
    # Identify causal variants in the results
    name_col <- if ("Name" %in% colnames(snp_res)) "Name" else "SNP"
    causal_snps <- bim$SNP[causal_idx]
    is_causal <- snp_res[[name_col]] %in% causal_snps
    
    # Plot all PIPs in gray
    plot(pip, type = "h", xlab = "SNP", ylab = "Posterior inclusion probability",
         col = "gray", lwd = 1)
    
    # Overlay causal variants in red
    points(which(is_causal), pip[is_causal], col = "red", pch = 19, cex = 1.2)
    segments(which(is_causal), 0, which(is_causal), pip[is_causal], 
             col = "red", lwd = 2)
    
    legend("topright", legend = c("Non-causal", "Causal"), 
           col = c("gray", "red"), lwd = c(1, 2), pch = c(NA, 19))
  } else {
    message("PIP column not found in snpRes file. Available columns: ", 
            paste(colnames(snp_res), collapse = ", "))
  }
}
```

# Run SBayesS using GWAS summary statistics

```{r}
# Summary-data-based analysis
system2(gctb, c(
  "--gwas-summary", sumstats_path,
  "--ldm-eigen", ldm_path,
  "--sbayes", "S",
  "--out", file.path(result_dir, "sim_sumstats")
))
```

# Inspect SBayesS results

```{r}
snp_res_sumstats_path <- file.path(result_dir, "sim_sumstats.snpRes")
par_res_sumstats_path <- file.path(result_dir, "sim_sumstats.parRes")

if (file.exists(snp_res_sumstats_path)) {
  snp_res_sumstats <- read.table(snp_res_sumstats_path, header = TRUE)
  cat("SBayesS SNP results:\n")
  head(snp_res_sumstats)
} else {
  message("Run SBayesS first to create ", snp_res_sumstats_path)
}

if (file.exists(par_res_sumstats_path)) {
  par_res_sumstats <- read.table(par_res_sumstats_path, header = TRUE)
  cat("\nSBayesS posterior estimates of model parameters:\n")
  print(par_res_sumstats)
} else {
  message("Run SBayesS first to create ", par_res_sumstats_path)
}

# Plot PIP values for SBayesS
if (file.exists(snp_res_sumstats_path)) {
  # Read results
  snp_res_sumstats <- read.table(snp_res_sumstats_path, header = TRUE)
  
  # Find the PIP column
  pip_col <- NULL
  for (col in c("PIP", "pip")) {
    if (col %in% colnames(snp_res_sumstats)) {
      pip_col <- col
      break
    }
  }
  
  if (!is.null(pip_col)) {
    pip <- snp_res_sumstats[[pip_col]]
    
    # Identify causal variants in the results
    name_col <- if ("Name" %in% colnames(snp_res_sumstats)) "Name" else "SNP"
    causal_snps <- bim$SNP[causal_idx]
    is_causal <- snp_res_sumstats[[name_col]] %in% causal_snps
    
    # Plot all PIPs in gray
    plot(pip, type = "h", xlab = "SNP", ylab = "Posterior inclusion probability",
         col = "gray", lwd = 1, main = "SBayesS PIP Values")
    
    # Overlay causal variants in red
    points(which(is_causal), pip[is_causal], col = "red", pch = 19, cex = 1.2)
    segments(which(is_causal), 0, which(is_causal), pip[is_causal], 
             col = "red", lwd = 2)
    
    legend("topright", legend = c("Non-causal", "Causal"), 
           col = c("gray", "red"), lwd = c(1, 2), pch = c(NA, 19))
  } else {
    message("PIP column not found in snpRes file. Available columns: ", 
            paste(colnames(snp_res_sumstats), collapse = ", "))
  }
}
```
